Matching Bullets

### 2015-12-15 imaging-paper.Rnw
Edits from Dr. Carriquiry

% The first character should be within \initial{}
\initial{F}\textbf{orensic examiners try to determine whether two guns were fired from the same barrel by analyzing them under a microscope in controlled conditions. Can we make the task simpler?}

@@

Bullet matching is a process in forensic science done to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of the conviction of criminals in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\cite{giannelli:2011}.

The new scrutiny of the procedure culminated with the 2009 National Academy of Sciences report~\cite{NAS:2009}. The report concluded that minimal research had been done exploring the reliability of these ballistic identification methods. In fact, the report says the forensic evidence ``including, for example, bite marks and firearm and toolmark identification is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline."

Rifling in the barrel creates striation marks on the bullet during the firing process. These marks are claimed to be unique to the barrel, as described in a 1992 AFTE article~\cite{afte:1992}. ``The theory of identification as it pertains to the comparison of toolmarks enables opinions of common origin to be made when the \emph{unique surface contours} of two toolmarks are in sufficient agreement" The article goes on to state that ``Significance is determined by the comparative examination of two or more sets of surface contour patterns comprised of individual peaks, ridges and furrows."

The primary goal in this type of forensic analysis is to identify these striation markings in an attempt to match two bullets. Since gun barrels should produce similar markings on bullets fired consecutively, the careful analysis of these surface images can help us determine whether two bullets are a match. However, this process has traditionally required forensic experts to perform the analysis, and required physical examination of the bullets. We've produced a framework which, in software, allows for the analysis of the surface topologies, and the transcription of the markings into a 2D plotting framework which we believe makes identification of matches easier and more rooted in science for both the expert and the lay person.

\section{Methodology}
Throughout this paper, we will work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\cite{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce known test bullets and unknown bullets for comparison.

3D topographical images of each bullet were taken using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). x3p is a container format conforming to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. x3p was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government Firearm Forensics researchers whose aim is to establish best practices for researchers using metrology in the forensic sciences.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[b]{\linewidth}
\caption{View from the side.\label{fig:sidex3p}}{%
      \includegraphics[width=\linewidth]{images/sidex3p.png}
    }
\end{subfigure}    
\begin{subfigure}[b]{\linewidth}
    \caption{View from the top.\label{fig:topx3p}}{%
    \includegraphics[width=\linewidth]{images/topx3p.png}
    }
\end{subfigure}
\caption{An example of a scan of a bullet land from one groove to the next. }
\end{figure}
Each fired bullet is provided in form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". For notational simplicity, we will refer to a particular land of some bullet as Bullet-X-Y, where X is the bullet number, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}, which show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, or a feature that pertains to a particular class of firearms and is not unique at the individual bullet level.

An initial naive approach to analyzing the striation patterns leads us to fixing a particular value of a coordinate, and then producing a plot of the heights across values of the other coordinate. Figure \ref{fig:fixedX} gives a plot resulting from such a procedure. It can be seen that the global structure of the land dominates the appearance of the plot. The grooves can be clearly identified on the left and right side, and the curvature of the surface is the most visible feature in the middle.

@@

It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

@@

For this, assume that $n$ data points are given in form of data tuples $(x_1, y_1), (x_2, y_2), ..., (x_n, y_n)$ that are (approximately) located on a circle. We want to estimate the location of the center and radius of the best fitting circle using a least squares approach.

We minimize the following expression:
\begin{equation}\label{eq:circle}
D = \sum_{i=1}^n \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right)^2
\end{equation}
We do this by differentiating $D$ with respect to $r, a,$ and $b$:
let us assume that $x_i$ and $y_i$ are centralized (i.e. $\sum x_i = \sum_i y_i = 0$). Note, if they are not, make a note of the current means, subtract them now and add them to $(\hat{a}, \hat{b})$ at the end.

\noindent
The  derivate of $D$ with respect to $r$ is:
\begin{eqnarray*}
\frac{d}{dr} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 r = \\
&=& 4 r \left( n r^2 - \sum_i (x_i-a)^2 - \sum_i(y_i-b)^2 \right) 
\end{eqnarray*}
For a minimum then holds:
\begin{equation}\label{eq:rmin}
\frac{d}{dr} D = 0 \stackrel{r \neq 0}{\iff} nr^2  = \sum_i (x_i-a)^2 + \sum_i(y_i-b)^2 
\end{equation}
%
%
The  derivative of $D$ with respect to $a$ is:
\begin{eqnarray*}
\frac{d}{da} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 (x_i - a) = \\
&=& -4 \left[ a \cdot nr^2 + \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]
\end{eqnarray*}
Using (\ref{eq:rmin}) for $nr^2$  in the equation above we get:
\begin{eqnarray*}
\frac{d}{da} D &=& -4 \left[  \sum_i a(x_i-a)^2 +  \sum_i a(y_i-b)^2  + \right. \\
&& \phantom{-4 \ \ } \left . \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]  = \\
&=& -4 \left[ \sum_i (x_i-a)^2 (a + x_i - a)  + \right.\\
&& \phantom{-4 \ \ } \left .\sum_i (x_i - a + a) (y_i - b)^2 \right] = \\
&=& -4 \left[ \sum_i (x_i-a)^2 x_i   + \sum_i x_i  (y_i - b)^2 \right] 
\stackrel{\begin{array}{c}\sum_i x_i = 0\\
\sum_i y_i = 0\end{array}}{=} \\
&=& -4 \left[ \sum_i x_i^3   + \sum_i x_i y_i^2  - 2a s_{xx} - 2b s_{xy} \right], 
\end{eqnarray*}
where $s_{xx} = \sum_i x_i^2, s_{xy} = \sum_i x_i y_i$ and $s_{yy} = \sum_i y_i^2$.

\noindent
Likewise, we get for the derivative of $D$ with respect to $b$:
\begin{eqnarray*}
\frac{d}{db} D &=& -4 \left[ \sum_i y_i^3   + \sum_i x_i^2 y_i - 2a s_{xy} - 2b s_{yy} \right] 
\end{eqnarray*}
For the minimum we therefore get a system of two linear equations in $a$ and $b$:
\begin{eqnarray*}
2 s_{xx} a + 2 s_{xy} b = c_1 && \text{ with } c_1 = \sum_i x_i^3 + x_i y_i^2 \\
2 s_{xy} a + 2 s_{yy} b = c_2 &&\text{ with } c_2 = \sum_i x_i^2 y_i + y_i^3.
\end{eqnarray*}
This resolves to 
\begin{eqnarray*}
\hat{a} &=& \frac{c_1 s_{yy} - c_2 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2},\\
\hat{b} &=& \frac{c_2 s_{xx} - c_1 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2}, \text{ and}\\
\hat{r^2} &=& \frac{1}{n}s_{xx} + \frac{1}{n}s_{yy} + \hat{a}^2 + \hat{b}^2.
\end{eqnarray*}

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$ and the land covers about \Sexpr{diff(range(dbr111$theta))} degrees.  Both of these estimates are consistent with a 9 mm bullet fired in a Ruger P-85.
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:residuala}Residual structure of a surface crosscut at $x = 1.5625$ (bottom of the bullet).}
<<residual2, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
qplot(data=subset(dbr111, x <= 1.5625), y, value-ypred, #colour=factor(x),
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  geom_hline(yintercept = 0, colour="grey50") +
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:residualb} Residual structure of a surface crosscut at $x = 100.00$}
<<residual, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
#qplot(data=subset(dbr111, x <= 80*1.5625^2 & x >=75*1.5625^2), y, value-ypred,
qplot(data=subset(dbr111, x == 100), y, value-ypred,  #colour=factor(x), 
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  geom_hline(yintercept = 0, colour="grey50") +
  theme_bw() + 
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}
\caption{\label{fig:residual} Residual structure of circular fits at two different cross-section. Both residual plots show systematic structures, indicating that a circular fit is not entirely appropriate.}
\end{figure}

A single cylinder as a fit is unlikely to be a particularly good fit, because there seem to be quite massive deformations in the vertical direction. Even when we fit a circle to each cross section of the bullet, as in Figure~\ref{fig:circlefits}, we can not address all of these issues. While the wider circumference at the base of the bullet can be resolved by individual circular fits, the systematic residual structure in Figure~\ref{fig:residualb} stays the same.

@@

\subsection{Loess Fit}
In an approach to improve upon the previous algorithm, we first have to eliminate the grooves from the images under the assumption that they do not contain relevant information for determining a match.
Fortunately, the location and appearance of the grooves in these 2d plots is quite consistent.
Surface measurements reach local maxima around the peak of the groove at either end of the range of $y$, we can then follow the descent of the surface measurements inwards to the valley of the groove. 
This marks the point at which we should trim the image. The procedure can be described as follows:

@@

\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step1}Step 1 the Loess fit algorithm. For a fixed crosscut ($x$ held constant at 243.75 $\mu m$) of surface measurements for bullet 1-5 are plotted across different values of $y$.}
<<loess_step1, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
br111_bullet <- get_bullet(paste(datadir,"Br1 Bullet 1-5.x3p", sep = "/"))
br121_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-1.x3p", sep = "/"))
br122_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-2.x3p", sep = "/"))
br123_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-3.x3p", sep = "/"))
br124_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-4.x3p", sep = "/"))
br125_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-5.x3p", sep = "/"))
br126_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-6.x3p", sep = "/"))

p1 <- qplot(y, value, data = br111_bullet, size=I(1)) + theme_bw() + ylab("Surface measurements")
p1
@
\end{subfigure}
%\begin{subfigure}[t]{\linewidth}
% \caption{\label{fig:loess_step2}Step 2 of the Loess fit algorithm. The height values are smoothed with a particular smoothing factor to remove noise in the deviations.}
<<loess_step2, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
subdata <- br111_bullet
smoothfactor <- 35

@@

p2 <- qplot(1.5625*(1:length(smoothed)), smoothed) + theme_bw() + xlab("y") +
  ylab("Surface measurements")
#p2
@
%\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step3}Step 2 of the Loess fit algorithm. The  height values are smoothed twice with a smoothing factor of 35. The orange rectangle shows an example of the smoothing window. Grooves and peaks will be detected, if they are not within the same window.}
<<loess_step3, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
xloc=250
p3 <- ggplot() +
  annotate("rect",xmin=xloc-17*1.5625, xmax=xloc+18*1.5625, ymin=80, ymax=200, fill=alpha("orange", alpha=0.5)) + 
  geom_point(aes(x=(1:length(smoothed_truefalse))*1.5625, y=smoothed_truefalse)) + theme_bw() +
    geom_vline(xintercept=xloc, colour="grey50") + xlab("y") + ylab("Surface measurements")

p3
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step47}Steps 3 -- 6 of the Loess fitting algorithm. After smoothing the $y$ values left and right minima are detected (marked by vertical lines, red indicating peaks and blue indicating valleys). Values outside these boundaries are removed (shown in grey)}
<<loess_step47, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
    lengthdiff <- length(subdata$value) - length(smoothed_truefalse)

@@

The ``smoothing factor" described in the algorithm represents the window size to use for a rolling average. Higher values of the smoothing factor therefore lead to more smoothed results. In the groove detection portion of this algorithm, we wish to remove most of the deviation but still maintain enough signal in the groove to detect it. Empirically, a value of 35 for the smoothing factor seems to work quite well (later in the manuscript we investigate varying the smoothing factor). It is important to note that the smoothing pass is done twice. That is, the smoothed data is once again smoothed by computing a new rolling average with the same smoothing factor. This bears some similarities to the ideas of John Tukey in his book Exploratory Data Analysis, where he describes a smoothing process called ``twicing" in which a second pass is made on the residuals computed from the first pass and then added back to the result~\cite{tukey:1977}. This has the effect of introducing a little bit more variance into the smoothed data. We instead performed a second smoothing pass on the smoothed data, which has the effect of weighting observations near the center highest, while the weights linearly drop off as you reach the end of the smoothing window or smoothing factor.

Next, we fit a Loess regression to the data. Figure \ref{fig:loess_fit} provides a look at the Loess fit, in blue, overlaid against the processed image of part 1 of bullet 1. The fit seems to do a reasonable job of capturing the structure of the image. Figure \ref{fig:loess_resid} shows the residuals from this fit. We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from bullet 1-5. This is shown in Figure~\ref{fig:loessfits}. Bullet 2-1 is still the best match for bullet 1-5, but now the details of the match are more pronounced.

@@

my.loess$fitted + ylab("Surface measurements")
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_resid} Residuals of Loess fit for bullet 1-5.}
<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE>>=
my.loess$resid + ylab("Residuals of Loess Fit")
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to bullet 1-5 (Barrel 1).}
\end{figure}

@@

In order to get a better understanding of  how  this matching works in known matches and non-matches, we next investigate the algorithm's performance in a test scenario.

>>>>

% The first character should be within \initial{}
\initial{F}\textbf{orensic examiners try to determine whether two bullets were fired from the same barrel by analyzing them under a microscope in controlled conditions. Can we make the task simpler?}

@@

Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of the conviction of criminals in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\cite{giannelli:2011}.

In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. ~\cite{NAS:2009}. The report concluded that minimal research had been done to explore the reliability of these ballistic identification methods. In fact, the report states that forensic evidence ``including, for example, bite marks and firearm and toolmark identification is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline."

Rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet during the firing process. These marks are assumed to be unique to the barrel, as described in a 1992 AFTE article~\cite{afte:1992}. ``The theory of identification as it pertains to the comparison of toolmarks enables opinions of common origin to be made when the \emph{unique surface contours} of two toolmarks are in sufficient agreement" The article goes on to state that ``Significance is determined by the comparative examination of two or more sets of surface contour patterns comprised of individual peaks, ridges and furrows."

The primary goal in this type of forensic analysis is to identify these striation markings in an attempt to match two bullets. Since gun barrels should produce similar markings on bullets fired consecutively, the careful analysis of these surface images can help us determine whether two bullets are a match. However, this process has traditionally required forensic experts to perform the analysis, and required physical examination of the bullets. We propose a framework which allows for the analysis of the surface topologies, and the transcription of the markings into a 2D plotting framework which we believe makes identification of matches easier and more rooted in science for both the expert and the lay person.

\section{Methodology}
Throughout this paper, we work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\cite{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce known test bullets and unknown bullets for comparison.

3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). x3p is a container format conforming to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. x3p was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in the forensic sciences.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[b]{\linewidth}
\caption{View from the side.\label{fig:sidex3p}}{%
      \includegraphics[width=\linewidth]{images/sidex3p.png}
    }
\end{subfigure}    
\begin{subfigure}[b]{\linewidth}
    \caption{View from the top.\label{fig:topx3p}}{%
    \includegraphics[width=\linewidth]{images/topx3p.png}
    }
\end{subfigure}
\caption{An example of a scan of a bullet land from one groove to the next. }
\end{figure}
Each fired bullet is provided in form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". For notational simplicity, we refer to a particular land of some bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, or a feature that pertains to a particular class of firearms and is not unique at the individual bullet level.

An initial naive approach to analyzing the striation patterns leads us to fixing a particular value of a coordinate, and then producing a plot of the heights across values of the other coordinate. Figure \ref{fig:fixedX} shows a plot resulting from such a procedure. It can be seen that the global structure of the land dominates the appearance of the plot. The grooves can be clearly identified on the left and right side, and the curvature of the surface is the most visible feature in the middle.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

@@

For this, assume that $n$ data points are given in the form of data tuples $(x_1, y_1), (x_2, y_2), ..., (x_n, y_n)$ that are (approximately) located on a circle. We want to estimate the location of the center and radius of the best fitting circle using a least squares approach.

We minimize the following expression:
\begin{equation}\label{eq:circle}
D = \sum_{i=1}^n \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right)^2,
\end{equation}
by differentiating $D$ with respect to $r, a,$ and $b$:
let us assume that $x_i$ and $y_i$ are centered (i.e. $\sum x_i = \sum_i y_i = 0$). Note, if they are not, make a note of the current means, subtract them now and add them to $(\hat{a}, \hat{b})$ at the end.

\noindent
The  derivate of $D$ with respect to $r$ is:
\begin{eqnarray*}
\frac{d}{dr} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 r = \\
&=& 4 r \left( n r^2 - \sum_i (x_i-a)^2 - \sum_i(y_i-b)^2 \right).
\end{eqnarray*}
At the minimum:
\begin{equation}\label{eq:rmin}
\frac{d}{dr} D = 0 \stackrel{r \neq 0}{\iff} nr^2  = \sum_i (x_i-a)^2 + \sum_i(y_i-b)^2.
\end{equation}
%
%
The  derivative of $D$ with respect to $a$ is:
\begin{eqnarray*}
\frac{d}{da} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 (x_i - a) = \\
&=& -4 \left[ a \cdot nr^2 + \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right].
\end{eqnarray*}
Using (\ref{eq:rmin}) for $nr^2$  in the equation above we get:
\begin{eqnarray*}
\frac{d}{da} D &=& -4 \left[  \sum_i a(x_i-a)^2 +  \sum_i a(y_i-b)^2  + \right. \\
&& \phantom{-4 \ \ } \left . \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]  = \\
&=& -4 \left[ \sum_i (x_i-a)^2 (a + x_i - a)  + \right.\\
&& \phantom{-4 \ \ } \left .\sum_i (x_i - a + a) (y_i - b)^2 \right] = \\
&=& -4 \left[ \sum_i (x_i-a)^2 x_i   + \sum_i x_i  (y_i - b)^2 \right] 
\stackrel{\begin{array}{c}\sum_i x_i = 0\\
\sum_i y_i = 0\end{array}}{=} \\
&=& -4 \left[ \sum_i x_i^3   + \sum_i x_i y_i^2  - 2a s_{xx} - 2b s_{xy} \right],
\end{eqnarray*}
where $s_{xx} = \sum_i x_i^2, s_{xy} = \sum_i x_i y_i$ and $s_{yy} = \sum_i y_i^2$.

\noindent
Likewise, we get for the derivative of $D$ with respect to $b$:
\begin{eqnarray*}
\frac{d}{db} D &=& -4 \left[ \sum_i y_i^3   + \sum_i x_i^2 y_i - 2a s_{xy} - 2b s_{yy} \right].
\end{eqnarray*}
To find the minimum we therefore get a system of two linear equations in $a$ and $b$:
\begin{eqnarray*}
2 s_{xx} a + 2 s_{xy} b = c_1 && \text{ with } c_1 = \sum_i x_i^3 + x_i y_i^2 \\
2 s_{xy} a + 2 s_{yy} b = c_2 &&\text{ with } c_2 = \sum_i x_i^2 y_i + y_i^3.
\end{eqnarray*}
The solution to the system is:
\begin{eqnarray*}
\hat{a} &=& \frac{c_1 s_{yy} - c_2 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2},\\
\hat{b} &=& \frac{c_2 s_{xx} - c_1 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2}, \text{ and}\\
\hat{r^2} &=& \frac{1}{n}s_{xx} + \frac{1}{n}s_{yy} + \hat{a}^2 + \hat{b}^2.
\end{eqnarray*}

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit.
In this instance, the radius is estimated as $\hat{r} = \Sexpr{pars$radius}\mu m = \Sexpr{pars$radius/1000}mm$ and the land covers about \Sexpr{diff(range(dbr111$theta))} degrees.  Both of these estimates are consistent with a 9 mm bullet fired by a Ruger P-85.
The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:residuala}Residual structure of a surface crosscut at $x = 1.5625$ (bottom of the bullet).}
<<residual2, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
qplot(data=subset(dbr111, x <= 1.5625), y, value-ypred, #colour=factor(x),
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  geom_hline(yintercept = 0, colour="grey50") +
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:residualb} Residual structure of a surface crosscut at $x = 100.00$}
<<residual, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
#qplot(data=subset(dbr111, x <= 80*1.5625^2 & x >=75*1.5625^2), y, value-ypred,
qplot(data=subset(dbr111, x == 100), y, value-ypred,  #colour=factor(x), 
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  geom_hline(yintercept = 0, colour="grey50") +
  theme_bw() + 
  ylab(expression(paste("Residuals (in ",mu,"m)", sep=""))) + 
  theme(legend.position="bottom")
@
\end{subfigure}
\caption{\label{fig:residual} Residual structure of circular fits at two different cross sections. Both residual plots show systematic structures, indicating that a circular fit is not entirely appropriate.}
\end{figure}

A single cylinder as a fit is unlikely to be a particularly good fit, because there seem to be quite massive deformations in the vertical direction. Even when we fit a circle to each cross section of the bullet, as in Figure~\ref{fig:circlefits}, this does not address all of these issues. While the wider circumference at the base of the bullet can be resolved by individual circular fits, the systematic residual structure in Figure~\ref{fig:residualb} stays the same.

@@

\subsection{Loess Fit}
In an attempt to improve upon the previous algorithm, we first have to eliminate the grooves from the images under the assumption that they do not contain relevant information for determining a match.
Fortunately, the location and appearance of the grooves in these 2d plots is quite consistent.
Surface measurements reach local maxima around the peak of the groove at either end of the range of $y$, we can then follow the descent of the surface measurements inwards to the valley of the groove. 
This marks the point at which we should trim the image. The procedure can be described as follows:

@@

\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step1}Step 1 the loess fit algorithm. For a fixed crosscut ($x$ held constant at 243.75 $\mu m$) of surface measurements for bullet 1-5 are plotted across different values of $y$.}
<<loess_step1, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
br111_bullet <- get_bullet(paste(datadir,"Br1 Bullet 1-5.x3p", sep = "/"))
br121_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-1.x3p", sep = "/"))
br122_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-2.x3p", sep = "/"))
br123_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-3.x3p", sep = "/"))
br124_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-4.x3p", sep = "/"))
br125_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-5.x3p", sep = "/"))
br126_bullet <- get_bullet(paste(datadir,"Br1 Bullet 2-6.x3p", sep = "/"))

p1 <- qplot(y, value, data = br111_bullet, size=I(1)) + theme_bw() + ylab("Surface measurements")
p1
@
\end{subfigure}
%\begin{subfigure}[t]{\linewidth}
% \caption{\label{fig:loess_step2}Step 2 of the loess fit algorithm. The height values are smoothed with a particular smoothing factor to remove noise in the deviations.}
<<loess_step2, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
subdata <- br111_bullet
smoothfactor <- 35

@@

p2 <- qplot(1.5625*(1:length(smoothed)), smoothed) + theme_bw() + xlab("y") +
  ylab("Surface measurements")
#p2
@
%\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step3}Step 2 of the loess fit algorithm. The  height values are smoothed twice with a smoothing factor of 35. The orange rectangle shows an example of the smoothing window. Grooves and peaks are detected, if they are not within the same window.}
<<loess_step3, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
xloc=250
p3 <- ggplot() +
  annotate("rect",xmin=xloc-17*1.5625, xmax=xloc+18*1.5625, ymin=80, ymax=200, fill=alpha("orange", alpha=0.5)) + 
  geom_point(aes(x=(1:length(smoothed_truefalse))*1.5625, y=smoothed_truefalse)) + theme_bw() +
    geom_vline(xintercept=xloc, colour="grey50") + xlab("y") + ylab("Surface measurements")

p3
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_step47}Steps 3 -- 6 of the loess fitting algorithm. After smoothing the $y$ values left and right minima are detected (marked by vertical lines, red indicating peaks and blue indicating valleys). Values outside these boundaries are removed (shown in grey)}
<<loess_step47, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE,  message=FALSE>>=
    lengthdiff <- length(subdata$value) - length(smoothed_truefalse)

@@

The ``smoothing factor" described in the algorithm represents the window size to use for a rolling average. Higher values of the smoothing factor therefore lead to more smoothed results. In the groove detection portion of this algorithm, we wish to remove most of the deviation but still maintain enough signal in the groove to detect it. Empirically, a value of 35 for the smoothing factor seems to work quite well (later in the manuscript we investigate varying the smoothing factor). It is important to note that the smoothing pass is done twice. That is, the smoothed data is once again smoothed by computing a new rolling average with the same smoothing factor. This bears some similarities to the ideas of John Tukey in his book Exploratory Data Analysis, where he describes a smoothing process called ``twicing" in which a second pass is made on the residuals computed from the first pass and then added back to the result~\cite{tukey:1977}. This has the effect of introducing a little bit more variance into the smoothed data. We instead performed a second smoothing pass on the smoothed data, which has the effect of weighting observations near the center highest, while the weights linearly drop off as we reach the end of the smoothing window or smoothing factor.

Next, we fit a loess regression to the data. Figure \ref{fig:loess_fit} provides a look at the loess fit, in blue, overlaid against the processed image of part 1 of bullet 1. The fit seems to do a reasonable job of capturing the structure of the image. Figure \ref{fig:loess_resid} shows the residuals from this fit. We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from bullet 1-5. This is shown in Figure~\ref{fig:loessfits}. Bullet 2-1 is still the best match for bullet 1-5, but now the details of the match are more pronounced.

@@

my.loess$fitted + ylab("Surface measurements")
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:loess_resid} Residuals of loess fit for bullet 1-5.}
<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\linewidth', warning=FALSE>>=
my.loess$resid + ylab("Residuals of loess Fit")
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to bullet 1-5 (Barrel 1).}
\end{figure}

@@

Determining a CMS threshold that needs to be exceeded for us to declare a match is beyond the scope of this work, even though it is critically important in practice. We provide some ideas in the next section. %In order to get a better understanding of  how  this matching works in known matches and non-matches, we next investigate the algorithm's performance in a test scenario.
